Applications of Many-Core Technologies to On-line Event Reconstruction in High Energy Physics Experiments

### 2013-11-14 papers/2013/NSS/NSS.tex
interim edits

% disable the running header on all pages following the title page
\pagestyle{empty}
%
\begin{document}
% paper title
\title{Applications of Many-Core Technologies to On-line Event Reconstruction in High Energy Physics
Experiments}
% Author list
\author{A.~Gianelle, 
  S.~Amerio, 
  D.~Bastieri, 
  M.~Corvo, 
  W.~Ketchum,
  T.~Liu, 
  A.~Lonardo, 
  D.~Lucchesi,
  S.~Poprocki, 
  R.~Rivera, 
  L.~Tosoratto,
  P.~Vicini
  and 
  P.~Wittich
\thanks{Manuscript received November 15, 2013.
This work was supported by the  US National Science Foundation, the US 
Department of Energy Office of Science and the Italian
Istituto Nazionale di Fisica Nucleare. This work was partially supported by the 
EU Framework Programme 7 project EURETILE under grant number 247846.}% <-this % stops a space
\thanks{D.~Bastieri and D.~Lucchesi are with University of Padova and INFN Padova.}% 
\thanks{A.~Gianelle is with INFN Padova.}% 
\thanks{M.~Corvo is with University of Ferrara.}% 
\thanks{S.~Amerio is with University of Padova}% 
\thanks{W.~Ketchum is with Los Alamos National Laboratory.}%
\thanks{T.~Liu and R.~Rivera are with Fermi National Accelerator Laboratory.}%
\thanks{A.~Lonardo, L.~Tosoratto and P.~Vicini are with INFN Roma.}% 
\thanks{S.~Poprocki and P.~Wittich are with Cornell University.}
}
% make the title area
\maketitle
% after \maketitle, add the following command to remove the header on the title page
\thispagestyle{empty}
% abstract usually in passive voice
\begin{abstract}
  Interest in parallel architectures applied to real time selections is 
  growing in High Energy Physics (HEP) experiments. In this paper we describe
  performance measurements of many-core devices 
 %Graphic Processing Units (GPUs) and Intel Many
  %Integrated Core architecture (MIC) 
  when applied to a typical HEP online task: 
  the selection of events based on the trajectories of charged particles.
  We use as benchmark a scaled-up version of the algorithm used at CDF 
  experiment at Tevatron for online track reconstruction - the SVT algorithm - 
  as a realistic  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment. We examine the
  complexity/performance trade-off in porting existing serial
  algorithms to many-core devices. We measure performance of
different architectures (Intel Xeon Phi and AMD GPU's, in addition 
to NVidia GPU's) and different software
environments (OpenCL, in addition to NVidia CUDA).
Measurements of both data processing  and data transfer latency are shown, 
considering different I/O strategies 
  to/from the parallel devices. \end{abstract}

\section{Introduction}
\IEEEPARstart{R}{eal}-time event reconstruction plays a fundamental role in High Energy
Physics (HEP) experiments at hadron colliders.  Reducing the rate of
data to be saved on tape from millions to hundreds of events per
second is critical. To increase the purity of the collected samples,
rate reduction has to be coupled with an initial selection of the most
interesting events.  In a typical hadron collider experiment, the
event rate has to be reduced from tens of MHz to a few kHz.  The
selection system (trigger) is usually organized in multiple levels,
each capable of performing a finer selection on more complex physics
objects describing the event. Trigger systems usually comprise a first
level based on custom hardware, followed by one or two levels usually
based on farms of general purpose processors.  At all levels, latency
is a concern: for a fixed processing time, the faster a decision is
rendered about accepting or rejecting an event improves the purity of
the collected data sample. The possibility of using commercial devices
at a low trigger level is very appealing: they are subject to
continuous performance improvements driven by the consumer market, are
less expensive than dedicated hardware, and are easier to support.
Among the commercial devices, many-core architectures such as Graphic
Processing Units (GPUs)~\cite{bib_gpu} and Intel Many Integrated Core
(MIC)~\cite{bib_intelMIC} are of particular interest for online
selections given their great computing power: the latest
\textsc{nvidia}~\cite{bib_nvidia} GPU architecture, Kepler, exceeds
Teraflop computing power. Moreover, high-level programming
architectures based on \textsc{c/c++} such as
\textsc{cuda}~\cite{bib_cuda} and \textsc{opencl}~\cite{bib_opencl}
make programming these devices more accessible to the general
physicist user.  The goal of this study is to investigate the
strengths and weaknesses of many-core devices when applied in a low
latency environment, with particular emphasis on the data transfer
latency to/from the device and the algorithm latency for processing on
the device in a manner similar to a typical HEP trigger application,
and to understand the cost/complexity ratio of porting legacy serial
code to many-core devices.

We showed initial studies on GPU performance in low-latency
environments ($\approx 100~\mu$s) in previous
papers~\cite{TIPP2011,NSS2012}.  In this paper we extend those studies
to include other many-core architectures (Intel MIC in addition to
GPUs).  The algorithm run on the parallel architecture is a complete
version of the fast track-fitting algorithm of the Silicon Vertex
Tracker (SVT) system at CDF~\cite{SVT1}.
We also investigate the performance of the algorithm on a GPU manufactured by AMD using a different
computing environment: \textsc{opencl}.
Starting with a serial algorithm
implemented on a CPU, we test an \textit{embarrassingly parallel}
algorithm on the Intel MIC environment. In this case each event is
handled independently by a core on the accelerator, and the
parallelization is achieved with only minor changes to the legacy code
and is only possible in the Intel MIC environment. Next we consider an
algorithm where we unroll three internal nested loops and run these in
parallel on a GPU, using the \textsc{cuda} environment. This second
approach is programmatically more complicated and less trivial to
implement. In neither case have we re-thought the basic algorithms or
the data structures used. To achieve optimal performance, these steps
would have to be taken.  As one might expect, the improvement from the
first approach is rather modest, and the second approach shows larger
performance gains. For GPUs, we also test different strategies to transfer
data to and from the device.

@@

To be run on NVidia GPUs, the code has been ported to \textsc{cuda}: each step --
unpack, combine and track fit -- is performed by a specific kernel; the
three nested loops are unrolled so that each GPU thread processes a
single combination of hits.  The \textsc{cuda} implementation makes use of 
\textsc{thrust}~\cite{bib_thrust}, a C++ template library for \textsc{cuda}, in the unpacking step.

@@

\begin{table*}[tbp]
  \caption{Capabilities of the many-core devices used in this study,
    according to the manufacturer's specifications. The first three
    are \textsc{nvidia} GPUs, the MIC 5110P is an Intel Xeon Phi, and the final one is an AMD GPU.}
  \label{tab_hwspecs}
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P & Radeon HD 7970 \\
    \hline
    \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 & 3790\\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320 & 264\\   
    Memory size (GB) & 3 & 5 & 6 & 8 & 3\\
    Number of cores & 448 & 2496 & 2688 & 240 & 2048\\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 & 1.375\\
    \hline
  \end{tabular}
\end{table*}

@@

We have an additional setup for testing the \textsc{opencl} implementation of the track 
fitting algorithm. On this setup, we use a 3.07 GHz Intel Core i7 CPU 950, which has four cores 
and up to eight computation threads. We can run the algorithm serially on this CPU (one core), or 
run the \textsc{opencl} algorithm and make use of the multi-core architecture. Additionally, we have an 
AMD Radeon HD 7970 GPU in a PCIe slot in this setup, on which we can also run the \textsc{opencl} algorithm.

\section{Results}
The input data consists of events with a fixed number of roads and
combinations: each event has 2048 combinations to be fitted. To
explore different data-taking conditions, the number of events ranges
from one to 3000, \emph{i.e.,} from 2048 to about six millions of combinations
to fit.

\subsection{Data processing}
Each data sample is processed 100 times by the track fitting algorithm. 
The average latency as a function of the number of fits is presented in 
Fig.~\ref{fig:algo_only_timing} for the serial,
embarrassingly parallel and parallel algorithms. We see that the
embarrassingly parallel algorithm gives a modest increase with respect
to the serial (CPU) algorithm. Switching to a fully parallel algorithm
affords a much more significant speed improvement. 
\begin{figure}[!t]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/TimeComp_MIC}
  \caption{Algorithm-only comparison for timing as a function of the
    number of track fits. We compare timing on CPUs (serial), Intel
    MIC (embarrassingly parallel), and GPUs (fully parallel), in
    blue, greed, and red, respectively. The GPUs exhibit the best
    performance due to the full parallelization. }
  \label{fig:algo_only_timing}
\end{figure}
\begin{figure}[!t]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/TimeCompZoom_MIC.png} 
  \caption{Algorithm-only comparison for timing as a function of the
    number of track fits: zoom in the low number of fits region. At
    low number of fits, the CPU performs better, due to start-up
    costs associated with data transfers to the accelerator card.}
  \label{fig:algo_only_timing_zoom}
\end{figure}
\begin{figure}[!t]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/Speedup_MIC}
  \caption{Speed-up with respect to a standard CPU (Intel Xeon
    E5630). The speed-ups plateau after about two million fits.}
  \label{fig:algo_only_speedup}
\end{figure}
Accelerators performance drop with decreasing number of fits, as can be seen 
in Fig.~\ref{fig:algo_only_timing_zoom}.
Figure~\ref{fig:algo_only_speedup} shows the speed-up with respect to the serial 
algorithm run on a standard CPU (Intel Xeon E5630): the maximum gain is 
obtained processing at least 500 events. This means that to fully exploit 
parallel architectures millions of fits have to be performed in parallel.

>>>>

% disable the running header on all pages following the title page
\pagestyle{empty}
%
\begin{document}
% paper title
\title{Applications of Many-Core Technologies to On-line Event Reconstruction in High Energy Physics
Experiments}
% Author list
\author{A.~Gianelle, 
  S.~Amerio, 
  D.~Bastieri, 
  M.~Corvo, 
  W.~Ketchum,
  T.~Liu, 
  A.~Lonardo, 
  D.~Lucchesi,
  S.~Poprocki, 
  R.~Rivera, 
  L.~Tosoratto,
  P.~Vicini
  and 
  P.~Wittich
\thanks{Manuscript received November 15, 2013.
This work was supported by the  US National Science Foundation, the US 
Department of Energy Office of Science and the Italian
Istituto Nazionale di Fisica Nucleare. This work was partially supported by the 
EU Framework Programme 7 project EURETILE under grant number 247846.}% <-this % stops a space
\thanks{D.~Bastieri and D.~Lucchesi are with University of Padova and INFN Padova.}% 
\thanks{A.~Gianelle is with INFN Padova.}% 
\thanks{M.~Corvo is with University of Ferrara.}% 
\thanks{S.~Amerio is with University of Padova}% 
\thanks{W.~Ketchum is with Los Alamos National Laboratory.}%
\thanks{T.~Liu and R.~Rivera are with Fermi National Accelerator Laboratory.}%
\thanks{A.~Lonardo, L.~Tosoratto and P.~Vicini are with INFN Roma.}% 
\thanks{S.~Poprocki and P.~Wittich are with Cornell University.}
}
% make the title area
\maketitle
% after \maketitle, add the following command to remove the header on the title page
\thispagestyle{empty}
% abstract usually in passive voice
\begin{abstract}
  Interest in many-core architectures applied to real time selections is 
  growing in High Energy Physics (HEP) experiments. In this paper we describe
  performance measurements of many-core devices 
  when applied to a typical HEP online task: 
  the selection of events based on the trajectories of charged particles.
  We use as benchmark a scaled-up version of the algorithm used at CDF 
  experiment at Tevatron for online track reconstruction - the SVT algorithm - 
  as a realistic  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment. We examine the
  complexity/performance trade-off in porting existing serial
  algorithms to many-core devices. We measure performance of
different architectures (Intel Xeon Phi and AMD GPUs, in addition 
to NVidia GPUs) and different software
environments (OpenCL, in addition to NVidia CUDA).
Measurements of both data processing  and data transfer latency are shown, 
considering different I/O strategies 
  to/from the many-core devices. \end{abstract}

\section{Introduction}
\IEEEPARstart{R}{eal}-time event reconstruction plays a fundamental role in High Energy
Physics (HEP) experiments at hadron colliders.  Reducing the rate of
data to be saved on tape is critical. To increase the purity of the collected samples,
rate reduction has to be coupled with an initial selection of the most
interesting events.  In a typical hadron collider experiment, the
event rate has to be reduced from tens of MHz to a few kHz.  The
selection system (trigger) is usually organized in successive levels,
each capable of performing a finer selection on more complex physics
objects describing the event. Trigger systems usually comprise a first
level based on custom hardware, followed by one or two levels usually
based on farms of general purpose processors.  At all levels, latency
is a concern: for a fixed processing time, the faster a decision is
rendered about accepting or rejecting an event improves the purity of
the collected data sample. The possibility of using commercial devices
at a low trigger level is very appealing: they are subject to
continuous performance improvements driven by the consumer market, are
less expensive than dedicated hardware, and are easier to support.
Among the commercial devices, many-core architectures such as Graphic
Processing Units (GPUs)~\cite{bib_gpu} and Intel Many Integrated Core
(MIC)~\cite{bib_intelMIC} are of particular interest for online
selections given their great computing power: the latest
\textsc{nvidia}~\cite{bib_nvidia} GPU architecture, Kepler, exceeds
Teraflop computing power. Moreover, high-level programming
architectures based on \textsc{c/c++} such as
\textsc{cuda}~\cite{bib_cuda} and \textsc{opencl}~\cite{bib_opencl}
make programming these devices more accessible to the general
physicist user.  The goal of this study is to investigate the
strengths and weaknesses of many-core devices when applied in a low
latency environment, with particular emphasis on the data transfer
latency to/from the device and the algorithm latency for processing on
the device in a manner similar to a typical HEP trigger application,
and to understand the cost/complexity ratio of porting legacy serial
code to many-core devices.

We showed initial studies on GPU performance in low-latency
environments ($\approx 100~\mu$s) in previous
papers~\cite{TIPP2011,NSS2012,CHEP2013}.  In this paper we extend
those studies to include other many-core architectures (Intel MIC in
addition to GPUs) and other programming toolsets (\textsc{opencl}).
The algorithm run on the parallel architecture is a complete version
of the fast track-fitting algorithm of the Silicon Vertex Tracker
(SVT) system at CDF~\cite{SVT1}.  We also investigate the performance
of the algorithm on a GPU manufactured by AMD using a different
computing environment: \textsc{opencl}.  Starting with a serial
algorithm implemented on a CPU, we test an \textit{embarrassingly
  parallel} algorithm on the Intel MIC environment. In this case each
event is handled independently by a core on the accelerator, and the
parallelization is achieved with only minor changes to the legacy
code.  This approach is only possible in the Intel MIC
environment. Next we consider an algorithm where we unroll three
internal nested loops and run these in parallel on a GPU, using the
\textsc{cuda} and \textsc{opencl} environments. This second approach
is programmatically more complicated and less trivial to implement. In
neither case have we re-thought the basic algorithms or the data
structures used. To achieve optimal performance, these steps would
have to be taken.  As one might expect, the improvement from the first
approach is rather modest, albeit easier to implement, and the second
approach shows larger performance gains. For GPUs, we also test
different strategies to transfer data to and from the device.

@@

To be run on NVidia GPUs, the code has been ported to \textsc{cuda}:
each step -- unpack, combine and track fit -- is performed by a
specific kernel; the three nested loops are unrolled so that each GPU
thread processes a single combination of hits.  The \textsc{cuda}
implementation makes use of \textsc{thrust}~\cite{bib_thrust}, a C++
template library for \textsc{cuda}, in the unpacking step. The
existence of template libraries such as \textsc{thrust} is an
advantage of the \textsc{cuda} environment.

@@

\begin{table*}[tbp]
  \caption{Capabilities of the many-core devices used in this study,
    according to the manufacturer's specifications. The first three
    are \textsc{nvidia} GPUs, the MIC 5110P is an Intel Xeon Phi, and
    the final one is an AMD GPU.  For Xeon Phi, the ``cores'' column
    counts the HW threads per core as equivalent to a GPU core.}
  \label{tab_hwspecs}
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P & Radeon HD 7970 \\
    \hline
     \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 & 3790\\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320 & 264\\   
    Memory size (GB) & 3 & 5 & 6 & 8 & 3\\
    Number of cores & 448 & 2496 & 2688 & 240 & 2048\\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 & 1.375\\
    \hline
  \end{tabular}
\end{table*}

@@

We have an additional setup for testing the \textsc{opencl}
implementation of the track fitting algorithm. Here, we use a 3.07 GHz
Intel Core i7 CPU 950, which has four cores and up to eight
computation threads. We run the algorithm serially on this CPU (one
core), and also run a CPU-based \textsc{opencl} algorithm which makes
use of the multi-core architecture. Additionally, we have an AMD
Radeon HD 7970 GPU in a PCIe slot in this setup, on which we also run
the \textsc{opencl} algorithm.

\section{Results}
The input data consists of events with a fixed number of roads and
combinations: each event has 2048 combinations to be fitted. To
explore different data-taking conditions, the number of events ranges
from one to 3000, \emph{i.e.,} between 2048 to about six millions of
combinations to fit.

\subsection{Data processing}
Each data sample is processed 100 times by the track fitting algorithm. 
The average latency as a function of the number of fits is presented in 
Fig.~\ref{fig:algo_only_timing} for the serial,
embarrassingly parallel and parallel algorithms. We see that the
embarrassingly parallel algorithm gives a modest increase with respect
to the serial (CPU) algorithm. Switching to a fully parallel algorithm
affords a much more significant speed improvement. 
\begin{figure}[!t]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/TimeComp_MIC}
  \caption{Algorithm-only comparison for timing as a function of the
    number of track fits. We compare timing on CPUs (serial), Intel
    MIC (embarrassingly parallel), and GPUs (fully parallel), in
    blue, greed, and red, respectively. The GPUs exhibit the best
    performance due to the full parallelization. }
  \label{fig:algo_only_timing}
\end{figure}
\begin{figure}[!t]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/TimeCompZoom_MIC.png} 
  \caption{Algorithm-only comparison for timing as a function of the
    number of track fits: zoom in the low number of fits region. At
    low number of fits, the CPU performs better, due to start-up
    costs associated with data transfers to the accelerator card.}
  \label{fig:algo_only_timing_zoom}
\end{figure}
\begin{figure}[!t]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/Speedup_MIC}
  \caption{Speed-up with respect to a standard CPU (Intel Xeon
    E5630). The speed-ups plateau after about two million fits.}
  \label{fig:algo_only_speedup}
\end{figure}
The accelerator card's performance drop with decreasing number of fits, as can be seen 
in Fig.~\ref{fig:algo_only_timing_zoom}, due to overhead.
Figure~\ref{fig:algo_only_speedup} shows the speed-up with respect to the serial 
algorithm run on a standard CPU (Intel Xeon E5630): the maximum gain is 
obtained processing at least 500 events. This means that to fully exploit 
parallel architectures millions of fits have to be performed in parallel.
